---
title: "Analyses of SAM-ratings"
author: "Maximillian Helles"
format: html
---

```{r setup}
#| include: false
source("_setup.R")
```

## SAM-ratings by Condition - Raw Means
```{r}
# Re-size dataframe by SAM Dimension
df_long <- df |>
  pivot_longer(
    cols = c(valence, arousal, regret),
    names_to = "sam_dimension",
    values_to = "rating"
  )

# Extract SAM-ratings per condition and get mean, sd, n, se. 
df_sam_summary <- df_long |>
  group_by(realized_condition, sam_dimension) |>
  summarise(
    mean_rating = mean(rating),
    sd_rating = sd(rating),
    n = n(),
    se = sd_rating / sqrt(n),
    .groups = "drop"
  )

# Display summaries in tibble
df_sam_summary

# Plot raw mean ratings
df_sam_summary |>
  ggplot(aes(x = realized_condition, y = mean_rating, fill = realized_condition)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_rating - se, ymax = mean_rating + se), width = 0.2) +
  facet_wrap(~ sam_dimension) +
  scale_fill_manual(values = c("Gain"= "#4C9F70","Relief" = "#8FB996", "FOMO" = "#E07A5F","Loss" = "#C44536")) +
  labs(x = NULL, y = "Raw Mean Ratings (±1 SE)") +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold")
  )
```
## Individual Responses and Means to SAM
```{r}
# Get individual raw means for each sam-rating per condition
df_participant <- df_long |>
  group_by(participant_id, realized_condition, sam_dimension) |>
  summarise(mean_rating = mean(rating), .groups = "drop")

# Plot individual raw means and responses
df_long |>
  ggplot(aes(x = realized_condition, y = rating, color = participant_id)) +
  geom_jitter(width = 0.2, height = 0.2, alpha = 0.5) +
  facet_wrap(~ sam_dimension) +
  theme_minimal() +
  geom_point(data = df_participant,
           aes(x = realized_condition, y = mean_rating, fill = participant_id),
           color = "black",         # fixed border color, outside aes()
           shape = 23,
           size = 3,
           stroke = 0.6) +
  facet_wrap(~ sam_dimension) +
  theme_minimal() + 
  scale_y_continuous(limits = c(1, 9), breaks = 1:9) +
  labs(x = NULL)
```
Individual responses are generally consistent with expectations / predictions. Two participants specifically reported ratings contrary to expectations:

- **Participant 007:** In both regret - and valence rating, 007 rated Gain ≈ FOMO & Relief ≈ Loss, indicating the participant failed / misunderstood the concept of positions (ASSET / CASH) and how that would affect their payout. 
- **Participant 005:** In Loss and FOMO, 005 ratings are elevated compared to other participants (except 007), suggesting a general positivity. They are, however, still monotonically decreasing. 

In arousal, two participants also reported lower ratings than the majority, but seems like it's just their own individual baselines.

## LMM Random Effects Comparison
Participants differ in baseline emotional response (captured by a random intercept) and may also differ in how strongly each condition shifts their ratings (captured by random slopes). Omitting a real source of between-participant variation deflates standard errors on the fixed effects and inflates Type-I error; including spurious variance components wastes degrees of freedom and can fail to converge.

Two specifications are compared per SAM dimension: a random intercept only, and uncorrelated random slopes for each non-reference condition. A fully correlated random-slope model was also considered but is not reported here: with n = 10 participants, the 6 covariance parameters were not identifiable and the model produced singular fits under all optimizers.

Even between the two retained specifications, the small participant pool matters. Each random-slope variance must be estimated from 10 per-participant deviations and may absorb noise rather than real structure, which is why BIC (penalizing complexity more heavily than AIC) sometimes favors the simpler intercept-only model. Selection proceeds per dimension based on the table below.

```{r}
df <- df |>
  mutate(
    cond_Relief = as.numeric(realized_condition == "Relief"),
    cond_FOMO   = as.numeric(realized_condition == "FOMO"),
    cond_Loss   = as.numeric(realized_condition == "Loss")
  )

# Fit three candidate random-effects structures for a given SAM dimension
# and return their fit statistics in a tidy row.
compare_models <- function(dv, data = df) {
  m_intercept <- lmer(
    as.formula(paste(dv, "~ realized_condition + (1 | participant_id)")),
    data = data
  )
  m_slopes_uncorr <- lmer(
    as.formula(paste(dv, "~ realized_condition +",
                     "(1 + cond_Relief + cond_FOMO + cond_Loss || participant_id)")),
    data = data
  )

  models <- list(
    "Random intercept"             = m_intercept,
    "Random slopes" = m_slopes_uncorr
  )

  tibble(
    Dimension = dv,
    Model     = names(models),
    Df        = vapply(models, \(m) attr(logLik(m), "df"), numeric(1)),
    logLik    = vapply(models, \(m) as.numeric(logLik(m)), numeric(1)),
    AIC       = vapply(models, AIC, numeric(1)),
    BIC       = vapply(models, BIC, numeric(1))
  )
}

model_comparison <- bind_rows(
  compare_models("arousal"),
  compare_models("regret"),
  compare_models("valence")
)

model_comparison
```
The random slopes model wins unambigously across all SAM dimensions, hence why the following will use a random slope LMM for modelling SAM-ratings.

## Fitting a LMM with random slope and intercept: Emotion ~ Condition
The Linear Mixed-Effects Model is determined by

emotion = β₀ + βs · conditions + u₀ + us · conditions + ε

where:
- β₀: fixed intercept, population-level mean of the reference condition (Gain).
- βs: fixed coefficients (β_Relief, β_FOMO, β_Loss), each the population-level shift from Gain.
- conditions: dummy-coded indicators (0/1) for each non-reference condition.
- u₀: random intercept, the participant's deviation from β₀.
- us: random slopes (u_Relief, u_FOMO, u_Loss), the participant's deviation from each fixed slope.
- ε: trial-level residual, drawn from N(0, σ²_ε) (noise).

The participant-level random effects (u₀, us) are jointly drawn from a multivariate normal N(0, Σ), where Σ is the random-effects covariance matrix. For this LMM, the *uncorrelated* version, in which Σ is diagonal and participants vary independently in baseline and in each condition effect will be used since initial testing revealed that the correlated version failed.

Since there is no universally correct way to determine the denominator degrees of freedom (ddf) for a t or F test in a general linear mixed model, p-values require approximation. To determine if the realized condition explains any variance at all, an ANOVA using the Kenward-Roger approximation is applied to the model. If so, 6 pairwise comparisons with multiple-comparison adjustment (tukey; studentized range distribution) will be performed.

#### Arousal
```{r}
model_arousal_condition <- lmer(
  arousal ~ realized_condition +
    (1 + cond_Relief + cond_FOMO + cond_Loss || participant_id),
  data = df
)
cat("---- ANOVA ----\n")
anova(model_arousal_condition, ddf = "Kenward-Roger")
cat("\n ---- PAIRWISE COMPARISONS ----\n")
emm_arousal <- emmeans(model_arousal_condition, pairwise ~ realized_condition, adjust = "tukey")
emm_arousal
```
All insignificant. LMM assumes the residuals and random effects are approximately Gaussian (follow a normal distribution), so to verify:

```{r}
qqnorm(resid(model_arousal_condition)); qqline(resid(model_arousal_condition))
ranef(model_arousal_condition)$participant_id 
```
QQ-plot looks fine, deviating slightly at the tails but nothing concerning. 
As the plot of individuals responses also suggested, participants 008 and 116 are ≈-3 points below population intercept (β₀). 008 has a low baseline, but 116 does vary more.

#### Regret
```{r}
model_regret_condition <- lmer(
  regret ~ realized_condition +
    (1 + cond_Relief + cond_FOMO + cond_Loss || participant_id),
  data = df
)
cat("---- ANOVA ----\n")
anova(model_regret_condition, ddf = "Kenward-Roger")
cat("\n ---- PAIRWISE COMPARISONS ----\n")
emm_regret <- emmeans(model_regret_condition, pairwise ~ realized_condition, adjust = "tukey")
emm_regret
```
Most adjusted p-values are significant except two: Gain - Relief (p = 0.13) and FOMO - Loss (p=0.077). Most likely due to low statistical power (n = 10). Checking effect sizes via cohen's d:
```{r}
eff_size(emm_regret, sigma = sigma(model_regret_condition), edf = df.residual(model_regret_condition)) |>
  suppressMessages()
```
Moderate-to-huge (ranging from -2.87 to -0.60) effects sizes across the board consistent with the plot of raw group means.

```{r}
qqnorm(resid(model_regret_condition)); qqline(resid(model_regret_condition))
ranef(model_regret_condition)$participant_id
```
QQ-plot looks OK again, though with slightly more deviating tails. Subject 999 differs most from fixed intercept at 0.89 but range is far narrower than arousal at [-0.54;0.89]. As seen in the plot of individual ratings, 999 nearly rates FOMO the same as Loss.

#### Valence
```{r}
model_valence_condition <- lmer(
  valence ~ realized_condition +
    (1 + cond_Relief + cond_FOMO + cond_Loss || participant_id),
  data = df
)
cat("---- ANOVA ----\n")
anova(model_valence_condition, ddf = "Kenward-Roger")
cat("\n---- PAIRWISE COMPARISONS ----\n")
emm_valence <- emmeans(model_valence_condition, pairwise ~ realized_condition, adjust = "tukey")
emm_valence
```
Again, lack of statistical power causes some contrasts to be insignificant. Checking effect sizes via cohen's d:
```{r}
eff_size(emm_valence, sigma = sigma(model_valence_condition), edf = df.residual(model_valence_condition)) |>
  suppressMessages()
```
Large-to-huge effect sizes on valence ratings across the board. Consistent with the raw means plot. Furtermore implying need for increased statistical power.

```{r}
qqnorm(resid(model_valence_condition)); qqline(resid(model_valence_condition))
ranef(model_valence_condition)$participant_id
```
As in Arousal and Regret, the QQ-plot show that residuals seem close to gaussian.

### Model Means Plot

```{r}
model_summary <- bind_rows(
  mutate(as.data.frame(emm_arousal$emmeans), sam_dimension = "arousal"),
  mutate(as.data.frame(emm_regret$emmeans),  sam_dimension = "regret"),
  mutate(as.data.frame(emm_valence$emmeans), sam_dimension = "valence")
) |>
  rename(mean_rating = emmean, se = SE)

model_summary |>
  ggplot(aes(x = realized_condition, y = mean_rating, fill = realized_condition)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_rating - se,
                    ymax = mean_rating + se),
                width = 0.2) +
  facet_wrap(~ sam_dimension) +
  scale_fill_manual(values = c("Gain"   = "#4C9F70",
                                "Relief" = "#8FB996",
                                "FOMO"   = "#E07A5F",
                                "Loss"   = "#C44536")) +
  labs(
    x        = NULL,
    y        = "Model-estimated Mean Ratings (±1 SE)",
  ) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold")
  )
```
## Fitting a LMM: Arousal ~ Invested (at spike)
```{r}
model_arousal_invested <- lmer(arousal ~ pos_at_jump + (1 | participant_id), 
                                data = df)

summary(model_arousal_invested)
```
Arousal tracks investment status (whether the participant held a position when the spike occurred), not the specific counterfactual structure (Gain vs. Loss, FOMO vs. Relief) that drives regret and valence.

## Jump and SAM-Rating 
```{r}
df |>
  ggplot(aes(x = abs(jump_pct), y = arousal, color = realized_condition)) +
  geom_jitter(alpha = 0.4, height = 0.2) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c("Gain"   = "#4C9F70",
                                 "Relief" = "#8FB996",
                                 "FOMO"   = "#E07A5F",
                                 "Loss"   = "#C44536")) +
  labs(x = "Jump magnitude (|jump_pct|)", 
       y = "Arousal", 
       color = "Condition") +
  theme_minimal()

df |>
  ggplot(aes(x = abs(jump_pct), y = regret, color = realized_condition)) +
  geom_jitter(alpha = 0.4, height = 0.2) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c("Gain"   = "#4C9F70",
                                 "Relief" = "#8FB996",
                                 "FOMO"   = "#E07A5F",
                                 "Loss"   = "#C44536")) +
  labs(x = "Jump magnitude (|jump_pct|)", 
       y = "Regret", 
       color = "Condition") +
  theme_minimal()

df |>
  ggplot(aes(x = abs(jump_pct), y = valence, color = realized_condition)) +
  geom_jitter(alpha = 0.4, height = 0.2) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = c("Gain"   = "#4C9F70",
                                 "Relief" = "#8FB996",
                                 "FOMO"   = "#E07A5F",
                                 "Loss"   = "#C44536")) +
  labs(x = "Jump magnitude (|jump_pct|)", 
       y = "Valence", 
       color = "Condition") +
  theme_minimal()
```
```{r}
model_arousal_jump <- lmer(arousal ~ abs(jump_pct) * realized_condition + 
                            (1 | participant_id), data = df)
anova(model_arousal_jump, ddf = "Kenward-Roger")

model_regret_jump <- lmer(regret ~ abs(jump_pct) * realized_condition + 
                            (1 | participant_id), data = df)
anova(model_regret_jump, ddf = "Kenward-Roger")

model_valence_jump <- lmer(valence ~ abs(jump_pct) * realized_condition + 
                            (1 | participant_id), data = df)
anova(model_valence_jump, ddf = "Kenward-Roger")
```
```{r}
emtrends(model_regret_jump,  ~ realized_condition, var = "abs(jump_pct)")
emtrends(model_valence_jump, ~ realized_condition, var = "abs(jump_pct)")
```

```{r}
df |>
  filter(realized_condition %in% c("Relief", "FOMO")) |>
  pivot_longer(c(regret, valence), names_to = "sam_dimension", values_to = "rating") |>
  ggplot(aes(x = abs(jump_pct), y = rating, color = realized_condition)) +
  geom_jitter(alpha = 0.3, height = 0.2) +
  geom_smooth(method = "lm") +
  facet_wrap(~ sam_dimension) +
  scale_color_manual(values = c("Relief" = "#8FB996", "FOMO" = "#E07A5F")) +
  labs(x = "Jump magnitude (|jump_pct|)", y = "SAM rating", color = "Condition") +
  theme_minimal()
```
An exploratory analysis examined whether jump magnitude (|jump_pct|) modulated SAM ratings beyond the condition effect. For arousal, neither the main effect nor the interaction with condition reached significance (all p > .3). For regret and valence, however, significant interactions emerged (regret: F(3, 773) = 3.17, p = .024; valence: F(3, 772) = 4.20, p = .006). Follow-up tests of per-condition slopes revealed that the magnitude effect was driven exclusively by the bystander conditions: in Relief trials, larger dodged negative jumps were associated with lower regret (slope = -41.4, 95% CI [-72.5, -10.4]) and higher valence (slope = +40.6, [+12.2, +69.1]); in FOMO trials, larger missed positive jumps were associated with lower valence (slope = -25.3, [-48.5, -2.1]). Invested conditions (Gain, Loss) showed no reliable magnitude effects, likely reflecting ceiling/floor effects in their SAM ratings.

## Capital & SAM-rating
```{r}
df |>
  pivot_longer(c(valence, arousal, regret), 
               names_to = "sam_dimension", values_to = "rating") |>
  group_by(block_id, realized_condition, sam_dimension) |>
  summarise(mean_rating = mean(rating, na.rm = TRUE),
            se = sd(rating, na.rm = TRUE) / sqrt(n()),
            .groups = "drop") |>
  ggplot(aes(x = realized_condition, y = mean_rating, fill = block_id)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = mean_rating - se, ymax = mean_rating + se),
                position = position_dodge(width = 0.8), width = 0.2) +
  facet_wrap(~ sam_dimension) +
  labs(x = NULL, y = "Mean rating (±1 SE)", fill = "Block") +
  theme_minimal()
```
```{r}
model_arousal_block <- lmer(arousal ~ realized_condition * block_id + 
                              (1 | participant_id), data = df)
anova(model_arousal_block, ddf = "Kenward-Roger")
emmeans(model_arousal_block, ~ block_id)


model_regret_block <- lmer(regret ~ realized_condition * block_id + 
                              (1 | participant_id), data = df)
anova(model_regret_block, ddf = "Kenward-Roger")

model_valence_block <- lmer(valence ~ realized_condition * block_id + 
                              (1 | participant_id), data = df)
anova(model_valence_block, ddf = "Kenward-Roger")
emmeans(model_valence_block, ~ realized_condition | block_id)
```
For valence, a significant condition-by-block interaction emerged, F(6, 767) = 2.77, p = .011, but the pattern was small and non-monotonic, with no clear evidence of linear amplification by capital. The condition effect remained the overwhelming driver of valence ratings in all blocks.

## Action proximity to jump vs. SAM-ratings
```{r}
df |>
  summarise(
    total      = n(),
    acted      = sum(!is.na(diff_action_jmp)),
    not_acted  = sum(is.na(diff_action_jmp))
  )

df |>
  filter(!is.na(diff_action_jmp)) |>
  mutate(abs_diff = abs(diff_action_jmp)) |>
  pivot_longer(c(valence, arousal, regret), 
               names_to = "sam_dimension", values_to = "rating") |>
  ggplot(aes(x = abs_diff, y = rating, color = realized_condition)) +
  geom_jitter(alpha = 0.3, height = 0.2) +
  geom_smooth(method = "lm") +
  facet_wrap(~ sam_dimension) +
  scale_color_manual(values = c("Gain"   = "#4C9F70",
                                 "Relief" = "#8FB996",
                                 "FOMO"   = "#E07A5F",
                                 "Loss"   = "#C44536")) +
  labs(x = "Action proximity to spike (|trials|)",
       y = "SAM rating",
       color = "Condition") +
  theme_minimal()
```


```{r}
df_acted <- df |>
  filter(!is.na(diff_action_jmp)) |>
  mutate(
    abs_diff = abs(diff_action_jmp),
    action_timing = if_else(diff_action_jmp < 0, "before_spike", "after_spike"),
    action_timing = factor(action_timing, levels = c("after_spike", "before_spike"))
  )

model_arousal_proximity <- lmer(arousal ~ abs_diff * realized_condition + 
                                 (1 | participant_id), data = df_acted)
anova(model_arousal_proximity, ddf = "Kenward-Roger")

model_regret_proximity <- lmer(regret ~ abs_diff * realized_condition + 
                                 (1 | participant_id), 
                                 data = df_acted)
anova(model_regret_proximity, ddf = "Kenward-Roger")

emtrends(model_regret_proximity, ~ realized_condition, var = "abs_diff")

model_valence_proximity <- lmer(valence ~ abs_diff * realized_condition + 
                                 (1 | participant_id), data = df_acted)
anova(model_valence_proximity, ddf = "Kenward-Roger")
```

