---
title: "Analyses of SAM-ratings"
author: "Maximillian Helles"
format: html
---

```{r setup}
#| include: false
source("_setup.R")
```

## SAM-ratings by Condition - Raw Means
```{r}
df_long <- df |>
  pivot_longer(
    cols      = c(valence, arousal, regret),
    names_to  = "sam_dimension",
    values_to = "rating"
  )

df_sam_summary <- df_long |>
  group_by(realized_condition, sam_dimension) |>
  summarise(
    mean_rating = mean(rating, na.rm = TRUE),
    sd_rating   = sd(rating,   na.rm = TRUE),
    n           = n(),
    se          = sd_rating / sqrt(n),
    .groups     = "drop"
  )

df_sam_summary

df_sam_summary |>
  ggplot(aes(x = realized_condition, y = mean_rating, fill = realized_condition)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_rating - se,
                    ymax = mean_rating + se),
                width = 0.2) +
  facet_wrap(~ sam_dimension) +
  scale_fill_manual(values = c("Gain"   = "#4C9F70",
                                "Relief" = "#8FB996",
                                "FOMO"   = "#E07A5F",
                                "Loss"   = "#C44536")) +
  labs(
    x        = NULL,
    y        = "Raw Mean Ratings (±1 SE)",
  ) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold")
  )
```
## Individual Responses and Means to SAM
```{r}
df_participant <- df_long |>
  group_by(participant_id, realized_condition, sam_dimension) |>
  summarise(mean_rating = mean(rating, na.rm = TRUE), .groups = "drop")

df_long |>
  ggplot(aes(x = realized_condition, y = rating, color = participant_id)) +
  geom_jitter(width = 0.2, height = 0.2, alpha = 0.5) +
  facet_wrap(~ sam_dimension) +
  theme_minimal() +
  geom_point(data = df_participant,
           aes(x = realized_condition, y = mean_rating, fill = participant_id),
           color = "black",         # fixed border color, outside aes()
           shape = 23,
           size = 3,
           stroke = 0.6) +
  facet_wrap(~ sam_dimension) +
  theme_minimal() + 
  scale_y_continuous(limits = c(1, 9), breaks = 1:9) +
  labs(x = NULL)
```
Individual responses generally consistent with expectation / prediction. 

#### Participant 007
In both regret and valence, participant 007 (color: sky blue) rated Gain ≈ FOMO & Relief ≈ Loss, indicating the participant failed / misunderstood the concept of positions (ASSET / CASH). This data will, however, still be used.

## Fitting a LMM: Emotion ~ Condition
#### Arousal
```{r}
model_arousal <- lmer(arousal ~ realized_condition + (1 | participant_id), data=df)
cat("---- LMM ----\n")
summary(model_arousal)
cat("\n ---- ANOVA ----\n")
anova(model_arousal, ddf = "Kenward-Roger")
cat("\n ---- PAIRWISE COMPARISONS ----\n")
emm_arousal <- emmeans(model_arousal, pairwise ~ realized_condition, adjust = "tukey")
emm_arousal
```
Residuals Diagnostic:
```{r}
qqnorm(resid(model_arousal)); qqline(resid(model_arousal))
ranef(model_arousal)$participant_id
```

#### Regret
```{r}
model_regret <- lmer(regret ~ realized_condition + (1 | participant_id), data=df)
cat("---- LMM ----\n")
summary(model_regret)
cat("\n ---- ANOVA ----\n")
anova(model_regret, ddf = "Kenward-Roger")
cat("\n ---- PAIRWISE COMPARISONS ----\n")
emm_regret <- emmeans(model_regret, pairwise ~ realized_condition, adjust = "tukey")
emm_regret
```
```{r}
qqnorm(resid(model_regret)); qqline(resid(model_regret))
ranef(model_regret)$participant_id
```


#### Valence
```{r}
model_valence <- lmer(valence ~ realized_condition + (1 | participant_id), data=df)
cat("---- LMM ----\n")
summary(model_valence)
cat("\n ---- ANOVA ----\n")
anova(model_valence, ddf = "Kenward-Roger")
cat("\n ---- PAIRWISE COMPARISONS ----\n")
emm_valence <- emmeans(model_valence, pairwise ~ realized_condition, adjust = "tukey")
emm_valence
```

```{r}
qqnorm(resid(model_regret)); qqline(resid(model_regret))
ranef(model_regret)$participant_id
```

```{r}
model_summary <- bind_rows(
  mutate(as.data.frame(emm_arousal$emmeans), sam_dimension = "arousal"),
  mutate(as.data.frame(emm_regret$emmeans),  sam_dimension = "regret"),
  mutate(as.data.frame(emm_valence$emmeans), sam_dimension = "valence")
) |>
  rename(mean_rating = emmean, se = SE)

model_summary |>
  ggplot(aes(x = realized_condition, y = mean_rating, fill = realized_condition)) +
  geom_col() +
  geom_errorbar(aes(ymin = mean_rating - se,
                    ymax = mean_rating + se),
                width = 0.2) +
  facet_wrap(~ sam_dimension) +
  scale_fill_manual(values = c("Gain"   = "#4C9F70",
                                "Relief" = "#8FB996",
                                "FOMO"   = "#E07A5F",
                                "Loss"   = "#C44536")) +
  labs(
    x        = NULL,
    y        = "Model-estimated Mean Ratings (±1 SE)",
  ) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    strip.text = element_text(face = "bold")
  )
```
## Fitting a LMM: Arousal ~ Invested (at spike)
```{r}
model_arousal_invested <- lmer(arousal ~ invested + (1 | participant_id), 
                                data = df)

summary(model_arousal_invested)
```
Arousal tracks investment status (whether the participant held a position when the spike occurred), not the specific counterfactual structure (Gain vs. Loss, FOMO vs. Relief) that drives regret and valence.

